Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-05-26 doc/thesis/discussion.tex
Started working on Sally's revisions

@@

It is important to note that \software{netabc} takes a transmission tree as
input, rather than a viral phylogeny. In reality, true transmission trees are
not available and must be estimated; these estimates are often based on the
viral phylogeny. Although this has been demonstrated to be a fair
approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
practice~\autocite[e.g.][]{stadler2013uncovering}, the topologies of a viral
phylogeny and transmission tree can differ
significantly~\autocite{ypma2013relating, hall2015epidemic} due to within-host
evolution and the sampling process. We have left the estimation of a
transmission tree up to the user. In theory, it is possible to incorporate the
process by which a viral phylogeny is generated along with a transmission tree
into our method, for example by simulating within-host dynamics. Although this
may be an avenue for future extension, we felt that it would obscure the
primary purpose of this work, which is to study contact network parameters. In
addition, there are a number of different methods available for inferring
transmission trees~\autocite{didelot2014bayesian, ypma2012unravelling,
jombart2011reconstructing, cottam2008integrating, hall2015epidemic}, some of
which incorporate geographic and/or epidemiological data not accommodated by
our method. We therefore felt it would be best to allow researchers to use
their own preferred method of constructing a transmission tree.

>>>>

\label{sec:disc}

@@

It is important to note that \software{netabc} takes a transmission tree as
input, rather than a viral phylogeny. In reality, true transmission trees are
not available and must be estimated; these estimates are often based on the
viral phylogeny. Although this has been demonstrated to be a fair
approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
practice~\autocite[e.g.][]{stadler2013uncovering}, the topologies of a viral
phylogeny and transmission tree can differ
significantly~\autocite{ypma2013relating, hall2015epidemic} due to within-host
evolution and the sampling process. We have left the estimation of a
transmission tree up to the user. In theory, it is possible to incorporate the
process by which a viral phylogeny is generated along with a transmission tree
into our method, for example by simulating within-host dynamics. {\color{blue}
\uline{Indeed, progress has very recently been demonstrated on this front in a
talk by \citeauthor{giardina2016inference}, who have independently developed in
a method similar to ours to fit contact network models to phylogenetic data
that additionally incorporates a within-host evolutionary model.}} Although
this may be an avenue for future extension, we felt that it would obscure the
primary purpose of this work, which is to study contact network parameters. In
addition, there are a number of different methods available for inferring
transmission trees~\autocite{didelot2014bayesian, ypma2012unravelling,
jombart2011reconstructing, cottam2008integrating, hall2015epidemic}, some of
which incorporate geographic and/or epidemiological data not accommodated by
our method. We therefore felt it would be best to allow researchers to use
their own preferred method of constructing a transmission tree.

### 2016-05-26 doc/thesis/intro.tex
Started working on Sally's revisions

A caveat of estimating transmission trees in this manner is that the
correspondence between the topologies of the viral phylogeny and transmission
tree is far from exact~\autocite{ypma2013relating, romero2014timing}.  Due to
intra-host diversity, the viral strain which is transmitted may have split from
another lineage within the donor long before the transmission event occurred.
Hence, the branching point in the viral phylogeny may be much earlier than that
in the transmission tree. Another possibility is that one host transmitted to
two or more recipients in one order, but the transmitted lineages originated
within the donor in a different order. In this case, the topology of the
transmission tree and the viral phylogeny will be mismatched. In practice, this
discordance has not proven an insurmountable problem: for example,
\textcite{leitner1996accurate, paraskevis2004phylogenetic} were able to
accurately recover a known transmission tree using a viral phylogeny. The
problem of accurately estimating transmission trees is an ongoing area of
research~\autocite{cottam2008integrating, jombart2011reconstructing,
ypma2012unravelling, morelli2012bayesian, didelot2014bayesian,
hall2015epidemic}. For example, \textcite{hall2015epidemic} developed a
Bayesian method to jointly estimate a transmission tree and viral phylogeny by
combining models of agent-based transmission, within-host population dynamics,
and sequence evolution.

@@

@@

To illustrate these concepts, consider the well-known linear model. For
clarity, we will restrict our attention to the case of one-dimensional data and
outcomes where $x = \set{x_1, \ldots, x_n}$ and $y = \set{y_1, \ldots, y_n}$
are vectors of real numbers. The linear model postulates that the outcomes are
linearly related to the data, modulo some noise introduced by measurement
error, environmental fluctuations, and other external factors. Formally, $y_i =
\beta x_i + \varepsilon_i$, where $\beta$ is the slope of the linear
relationship, and $\varepsilon_i$ is the error associated with measurement $i$.
We can make this model a statistical one by hypothesizing a distribution for
the error terms $\varepsilon_i$; most commonly, it is assumed that they are
normally distributed with variance $\sigma$. In mathematical terms, $Y_i \sim
\beta x_i + \N(0, \sigma^2)$, where ``$\sim$'' means ``is distributed as''. We
can see from this formulation that the model is parametric, with parameters
$\theta$ = ($\beta$, $\sigma$). Moreover, we can write down the probability
density $\pi$ of observing outcome $y_i$ given the parameters,
\[
  \pi(y \mid \beta, \sigma) = 
  \prod_{i=1}^n f_{\N(0, \sigma^2)} (y_i - \beta x_i),
\]
where $f_{\N(0, \sigma^2)}$ is the probability density of the normal
distribution with mean zero and variance $\sigma^2$. Note that we are treating
the $x_i$ as fixed quantities, and therefore have not conditioned the
probability density on $x$. Also, we have assumed that all the $y_i$ are
independent.

@@

Ordinarily, Bayesian inference targets the posterior distribution $f(\theta
\mid y)$. That is, in the Bayesian framework, model parameters with higher
posterior density are ``better'' in the sense that they offer a more credible
explanation for the observed data. Approximate Bayesian computation offers an
alternative metric for parameter credibility, namely the similarity of
simulated datasets to the observed data. If datasets simulated under the model
closely resemble the real data, it follows that the model is a reasonable
approximation to the real-world process generating the observed data. More
formally, suppose we have a distance measure $\rho$ defined on the space of all
possible data our model could generate. \gls{ABC} aims to sample from the joint
posterior distribution of model parameters and simulated datasets $z$ which are
within some small distance $\varepsilon$ of the observed data $y$,
\[
  \pi_{\varepsilon}(\theta, z \mid y) =
  \frac{\pi(\theta) f(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
  {\int_{A_{\varepsilon, y} \times \Theta} \pi(\theta) f(z \mid \theta) \d \theta}.
\]
Here, $A_{\varepsilon, y}$ is an $\varepsilon$-ball around $y$ with
respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
$\I$ is the indicator function~\autocite{marin2012approximate}. As we shall
see in the next section, this distribution can be sampled from exactly. The
word ``approximate'' derives from the assumption that, for a suitably chosen
distance $\rho$ and a small enough $\varepsilon$, the marginal in $z$ of
this distribution approximates the posterior of
interest~\autocite{marin2012approximate}. That is,
\[
  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx f(\theta \mid y).
\]
The distribution $\pi_\varepsilon(\theta, z \mid y)$ is variously referred to
as the \textit{\gls{ABC} target distribution} or the \gls{ABC} approximation
to the posterior. The intuition for why this approximation might be reasonable
comes from the fact that, when $\varepsilon = 0$ and $\rho$ satisfies $\rho(z,
y) = 0 \Rightarrow x = y$, the integral on the left is exactly equal to the
posterior. Thus, by taking $\varepsilon$ small, we should attain something
close to the posterior {\color{blue}\uline{if $\rho$ captures the similarity
between datasets reasonably well. However, the accuracy of the \gls{ABC}
approximation depends heavily on the choice of distance function. }}

In many applications {\color{purple}\uline{(eg. \autocite{fu1997estimating,
tanaka2006using})}}, $\rho$ is defined as $\rho(S(\cdot), S(\cdot))$ where $S$
is a function which maps data points into a vector of summary statistics.
{\color{blue}\uline{In the context of \gls{ABC}, a summary statistic $S$ is
called \defn{sufficient} if} 
\[
  f(\theta \mid y) = f(\theta \mid S(y)).
\]
\uline{That is, sufficiency implies that the data can be replaced with the
summary statistic without losing any information about the posterior
distribution~\autocite{marjoram2006modern}. For most problems, it is not
possible to find sufficient summary statistics~\autocite{marjoram2006modern}.
TODO
}}

This can be useful if the data are high-dimensional or of a complex type, but
it is not strictly necessary. For instance, if the data are numeric and of low
dimension, the distance function may simply be the Euclidean
distance~\autocite{sisson2007sequential}. \textcite{poon2015phylodynamic}
proposed the use of a kernel function (defined in \cref{subsec:treeshape}),
which allows

@@

%\subsection{Convergence properties}
%
%The convergence of \gls{ABC} to the true posterior distribution of interest
%depends on two components: the convergence of the underlying algorithm to the
%\gls{ABC} target distribution, and the resemblance of this target distribution
%to the true
%posterior~\autocite{marin2012approximate,fearnhead2012constructing}. We shall
%address these components one at a time. Rejection \gls{ABC} and
%\gls{ABC}-\gls{MCMC} use, respectively, rejection sampling and \gls{MCMC} to
%sample from the target distributions. Both of these methods are
%well-established and their convergence properties have been extensively
%studied. Since we do not apply rejection or \gls{MCMC} in this work, we shall
%restrict our attention to the convergence properties of \gls{SMC}.

>>>>

A caveat of estimating transmission trees in this manner is that the
correspondence between the topologies of the viral phylogeny and transmission
tree is far from exact~\autocite{ypma2013relating, romero2014timing}.  Due to
intra-host diversity, the viral strain which is transmitted may have split from
another lineage within the donor long before the transmission event occurred.
Hence, the branching point in the viral phylogeny may be much earlier than that
in the transmission tree. Another possibility is that one host transmitted to
two or more recipients in one order, but the transmitted lineages originated
within the donor in a different order. In this case, the topology of the
transmission tree and the viral phylogeny will be mismatched. In practice, this
discordance has not proven an insurmountable problem: for example,
\textcite{leitner1996accurate, paraskevis2004phylogenetic} were able to
accurately recover known transmission trees using viral phylogenies. The
problem of accurately estimating transmission trees is an ongoing area of
research~\autocite{cottam2008integrating, jombart2011reconstructing,
ypma2012unravelling, morelli2012bayesian, didelot2014bayesian,
hall2015epidemic}. For example, \textcite{hall2015epidemic} developed a
Bayesian method to jointly estimate a transmission tree and viral phylogeny by
combining models of agent-based transmission, within-host population dynamics,
and sequence evolution.

@@

{\color{blue}\uline{
Recall that the primary objective of our work is to develop a statistical
inference method for estimating contact network parameters from transmission
trees. In particular, for a network model with parameters $\theta$ and an input
transmission tree $T$, we are interested in the posterior distribution}
\[
  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}.
\]
\uline{As outlined in} \cref{sec:obj}, \uline{both the likelihood $\Pr(T \mid
\theta)$ and the normalizing constant $\Pr(T)$ are likely computationally
intractable. Hence, rather than computing the posterior distribution
analytically, we will attempt to approximate it using a \defn{Monte Carlo}
approach.
TODO
}}

@@

To illustrate these concepts, consider the well-known linear model. For
clarity, we will restrict our attention to the case of one-dimensional data and
outcomes where $x = \set{x_1, \ldots, x_n}$ and $y = \set{y_1, \ldots, y_n}$
are vectors of real numbers. The linear model postulates that the outcomes are
linearly related to the data, modulo some noise introduced by measurement
error, environmental fluctuations, and other external factors. Formally, $y_i =
\beta x_i + \varepsilon_i$, where $\beta$ is the slope of the linear
relationship, and $\varepsilon_i$ is the error associated with measurement $i$.
We can make this model a statistical one by hypothesizing a distribution for
the error terms $\varepsilon_i$; most commonly, it is assumed that they are
normally distributed with variance $\sigma$. In mathematical terms, $y_i \sim
\beta x_i + \N(0, \sigma^2)$, where ``$\sim$'' means ``is distributed as''. We
can see from this formulation that the model is parametric, with parameters
$\theta$ = ($\beta$, $\sigma$). Moreover, we can write down the probability
density $\pi$ of observing outcome $y_i$ given the parameters,
\[
  \pi(y \mid \beta, \sigma) = 
  \prod_{i=1}^n f_{\N(0, \sigma^2)} (y_i - \beta x_i),
\]
where $f_{\N(0, \sigma^2)}$ is the probability density of the normal
distribution with mean zero and variance $\sigma^2$. Note that we are treating
the $x_i$ as fixed quantities, and therefore have not conditioned the
probability density on $x$. Also, we have assumed that all the $y_i$ are
independent.

@@

Ordinarily, Bayesian inference targets the posterior distribution $f(\theta
\mid y)$. That is, in the Bayesian framework, model parameters with higher
posterior density are ``better'' in the sense that they offer a more credible
explanation for the observed data. Approximate Bayesian computation offers an
alternative metric for parameter credibility, namely the similarity of
simulated datasets to the observed data. If datasets simulated under the model
closely resemble the real data, it follows that the model is a reasonable
approximation to the real-world process generating the observed data. More
formally, suppose we have a distance measure $\rho$ defined on the space of all
possible data our model could generate. \gls{ABC} aims to sample from the joint
posterior distribution of model parameters and simulated datasets $z$ which are
within some small distance $\varepsilon$ of the observed data $y$,
\[
  \pi_{\varepsilon}(\theta, z \mid y) =
  \frac{\pi(\theta) f(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
  {\int_{A_{\varepsilon, y} \times \Theta} \pi(\theta) f(z \mid \theta) \d \theta}.
\]
Here, $A_{\varepsilon, y}$ is an $\varepsilon$-ball around $y$ with
respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
$\I$ is the indicator function~\autocite{marin2012approximate}. As we shall
see in the next section, this distribution can be sampled from exactly. The
word ``approximate'' derives from the assumption that, for a suitably chosen
distance $\rho$ and a small enough $\varepsilon$, the marginal in $z$ of
this distribution approximates the posterior of
interest~\autocite{marin2012approximate}. That is,
\[
  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx f(\theta \mid y).
\]
The distribution $\pi_\varepsilon(\theta, z \mid y)$ is variously referred to
as the \textit{\gls{ABC} target distribution} or the \gls{ABC} approximation
to the posterior. The intuition for why this approximation might be reasonable
comes from the fact that, when $\varepsilon = 0$, the $\varepsilon$-ball around
$y$ should contain only $y$ itself, hence the integral on the left is exactly
equal to the posterior. Thus, by taking $\varepsilon$ small, we should attain
something close to the posterior {\color{blue}\uline{if $\rho$ captures the
similarity between datasets reasonably well. However, the accuracy of the
\gls{ABC} approximation depends heavily on the choice of distance function
\autocite{aeschbacher2012novel,blum2013comparative}.}}

In many applications {\color{blue}\uline{(eg. \autocite{fu1997estimating,
tanaka2006using})}}, $\rho$ is defined as $\rho(S(\cdot), S(\cdot))$ where $S$
is a function which maps data points into a vector of summary statistics.
{\color{blue}\uline{In the context of \gls{ABC}, a summary statistic $S$ is
called \defn{sufficient} if} 
\[
  f(\theta \mid y) = f(\theta \mid S(y)).
\]
\uline{That is, sufficiency implies that the data can be replaced with the
summary statistic without losing any information about the posterior
distribution~\autocite{marjoram2006modern}. For most problems, it is not
possible to find sufficient summary statistics~\autocite{marjoram2006modern}.
A number of sophisticated methods have been developed for selecting and
weighting summary statistics based on various optimality criteria}
\autocite[][and references therein]{aeschbacher2012novel,
blum2013comparative}}.

Summary statistics can be useful if the data are high-dimensional or of a
complex type, but they are not strictly necessary. For instance, if the data
are numeric and of low dimension, the distance function may simply be the
Euclidean distance~\autocite{sisson2007sequential}.
{\color{blue}\textcite{park2015k2} \uline{proposed the use of a kernel function}
(defined in \cref{subsec:treeshape}) \uline{in place of a distance function.
The authors referred to their approach as ``double-kernel \gls{ABC}'' due to
the use of a second kernel function to compute the weights of the particles.
The work by \textcite{poon2015phylodynamic}, upon which ours is based, employed
a similar approach, replacing the likelihood ratio in Bayesian \gls{MCMC} with
a ratio of kernel scores.}}

@@

%\subsection{Convergence properties}
%
%The convergence of \gls{ABC} to the true posterior distribution of interest
%depends on two components: the convergence of the underlying algorithm to the
%\gls{ABC} target distribution, and the resemblance of this target distribution
%to the true
%posterior~\autocite{marin2012approximate,fearnhead2012constructing}. We shall
%address these components one at a time. Rejection \gls{ABC} and
%\gls{ABC}-\gls{MCMC} use, respectively, rejection sampling and \gls{MCMC} to
%sample from the target distributions. Both of these methods are
%well-established and their convergence properties have been extensively
%studied. Since we do not apply rejection or \gls{MCMC} in this work, we shall
%}restrict our attention to the convergence properties of \gls{SMC}.

\section{Summary}

{\color{blue}\uline{ Our method integrates the four distinct research areas
  just described: phylogenetics, contact networks, \acrlong{SMC}, and
  \acrlong{ABC}. The first two topics together form the problem domain.
  Phylogenetic data is the input to our method, while estimates of the
  parameters of contact network models are the desired output. The latter two
  topics define the algorithm and statistical framework that our inference
method will use. }}
